Resolvable Ambiguity

### 2019-05-30 paper/paper.tex
EOD

\begin{figure}
  \begin{tabular}{@{}ll@{}}
      Terminals & $t \in \T$ \\
      Non-terminals & $N \in \NT$ \\
      Labels & $l \in \Labels$ \\
      Marks & $m \subseteq \Labels$ \\
      Regular expressions & $r ::= t \mid N_m \mid r \cdot r \mid r + r \mid \epsilon \mid r^{*}$ \\
      Labelled productions & $N -> l : r$ \\
  \end{tabular}
  \caption{The abstract syntax of a language definition.}
  \label{fig:input-language-definition}
\end{figure}

\begin{figure}
  \begin{tabular}{@{}l@{\quad$->$\quad}c@{ $:$\quad}l@{}}
    $E$ & $l$ & \verb|'['| ($E$ (\verb|';'| $E$)$^{*}$ $+$ $\epsilon$) \verb|']'| \\
    $E$ & $a$ & $E$ \verb|'+'| $E$ \\
    $E$ & $m$ & $E_{\{a\}}$ \verb|'*'| $E_{\{a\}}$ \\
    $E$ & $n$ & $N$ \\
  \end{tabular}
  \caption{The input language definition used as a running example, an expression language with lists, addition, and multiplication, with precedence defined, but not associativity. Assumes that $N$ matches a numeric terminal.}
  \label{fig:running-example-definition}
\end{figure}

@@

\begin{figure}
  \begin{subfigure}[t]{.45\linewidth}
    \centering
    \begin{tabular}{@{}l@{\quad$->$\quad}l@{}}
      \toprule
      $E$ & \verb|'['| $E_{l1}$ \verb|']'| \\
      $E$ & $E$ \verb|'+'| $E$ \\
      $E$ & $E$ \verb|'*'| $E$ \\
      $E$ & $N$ \\
      \midrule
      $E_{l1}$ & $\epsilon$ \\
      $E_{l1}$ & $E$ $E_{l2}$ \\
      \midrule
      $E_{l2}$ & $\epsilon$ \\
      $E_{l2}$ & \verb|';'| $E$ $E_{l2}$ \\
      \bottomrule
    \end{tabular}
    \caption{$G_D$, the generated abstract syntax.}
    \label{fig:running-example-generated:w}
  \end{subfigure}%
%
  \begin{subfigure}[t]{.54\linewidth}
    \centering
    \begin{tabular}{@{}l@{\quad$->$\quad}c@{$($ }l@{}}
      \toprule
      $E$ & $l$ & \verb|'['| $(\epsilon + E($\verb|';'| $E)^{*})$ \verb|']'| $)$ \\
      $E$ & $a$ & $E$ \verb|'+'| $E$ $)$ \\
      $E$ & $m$ & $E$ \verb|'*'| $E$ $)$ \\
      $E$ & $n$ & $N$ $)$ \\
      \bottomrule
    \end{tabular}
    \caption{$T_D$, the parse trees of $G_D$.}
    \label{fig:running-example-generated:t}
  \end{subfigure}

  \begin{subfigure}[t]{.45\linewidth}
    \centering
    \begin{tabular}{@{}l@{\quad$->$\quad}l@{}}
      \toprule
      $E$ & \verb|'['| $E_{l1}$ \verb|']'| \\
      $E$ & $E$ \verb|'+'| $E$ \\
      $E$ & $E_{\{a\}}$ \verb|'*'| $E_{\{a\}}$ \\
      $E$ & $N$ \\
      $E$ & \verb|'('| $E$ \verb|')'| \\
      \midrule
      $E_{\{a\}}$ & \verb|'['| $E_{l1}$ \verb|']'| \\
      $E_{\{a\}}$ & $E_{\{a\}}$ \verb|'*'| $E_{\{a\}}$ \\
      $E_{\{a\}}$ & $N$ \\
      $E_{\{a\}}$ & \verb|'('| $E$ \verb|')'| \\
      \midrule
      $E_{l1}$ & $\epsilon$ \\
      $E_{l1}$ & $E$ $E_{l2}$ \\
      \midrule
      $E_{l2}$ & $\epsilon$ \\
      $E_{l2}$ & \verb|';'| $E$ $E_{l2}$ \\
      \bottomrule
    \end{tabular}
    \caption{$G'_D$, the generated concrete syntax.}
    \label{fig:running-example-generated:w-prime}
  \end{subfigure}%
%
  \begin{subfigure}[t]{.54\linewidth}
    \centering
    \begin{tabular}{@{}l@{\quad$->$\quad}c@{$($ }l@{}}
      \toprule
      $E$ & $l$ & \verb|'['| $(\epsilon + E($\verb|';'| $E)^{*})$ \verb|']'| $)$ \\
      $E$ & $a$ & $E$ \verb|'+'| $E$ $)$ \\
      $E$ & $m$ & $E_{\{a\}}$ \verb|'*'| $E_{\{a\}}$ $)$ \\
      $E$ & $n$ & $N$ $)$ \\
      $E$ & $g$ & \verb|'('| $E$ \verb|')'| $)$ \\
      \midrule
      $E_{\{a\}}$ & $l$ & \verb|'['| $(\epsilon + E($\verb|';'| $E)^{*})$ \verb|']'| $)$ \\
      $E_{\{a\}}$ & $m$ & $E_{\{a\}}$ \verb|'*'| $E_{\{a\}}$ $)$ \\
      $E_{\{a\}}$ & $n$ & $N$ $)$ \\
      $E_{\{a\}}$ & $g$ & \verb|'('| $E$ \verb|')'| $)$ \\
      \bottomrule
    \end{tabular}
    \caption{$T'_D$, the parse trees of $G'_D$.}
    \label{fig:running-example-generated:t-prime}
  \end{subfigure}%
  \caption{The generated grammars.}
  \label{fig:running-example-generated}
\end{figure}

@@

\begin{figure}
  \begin{subfigure}[b]{.40\linewidth}
    \[1 + 2 * 3\]
    \caption{Example word in $L(G_D)$.}
  \end{subfigure}
  \begin{subfigure}[b]{.55\linewidth}
    \begin{center}
    \Tree [.{$m$}
        [.{$a$}
          [.{$n$} \terminal{1} ]
          \terminal{+}
          [.{$n$} \terminal{2} ] ]
        \terminal{*}
        [.{$n$} \terminal{3} ] ]
    \end{center}
    \[m(a(n(\terminal{1}) \terminal{+} n(\terminal{2})) \terminal{*} n(\terminal{3}))\]
    \caption{Example tree in $L(T_D)$ and visualization.}
  \end{subfigure}

  \begin{subfigure}[b]{.40\linewidth}
    \[(1 + 2) * 3\]
    \caption{Example word in $L(G'_D)$.}
  \end{subfigure}
  \begin{subfigure}[b]{.55\linewidth}
    \begin{center}
    \Tree [.{$m$}
        [.{$g$}
          \terminal{(}
          [.{$a$}
            [.{$n$} \terminal{1} ]
            \terminal{+}
            [.{$n$} \terminal{2} ] ]
          \terminal{)} ]
        \terminal{*}
        [.{$n$} \terminal{3} ] ]
    \end{center}
    \[m(g( \terminal{(} a(n( \terminal{1} ) \terminal{+} n( \terminal{2} )) \terminal{)} ) \terminal{*} n( \terminal{3} ))\]
    \caption{Example tree in $L(T'_D)$ and visualization.}
  \end{subfigure}
  \caption{Example with elements from each generated language that correspond to each other. The leaf terminals in the tree languages appear underlined to distinguish the two kinds of parentheses.}
  \label{fig:example-words-in-square}
\end{figure}
}

@@

\begin{figure}
  \begin{tikzcd}
    L(G_D) & L(T_D) \arrow[l, "\yield"] \\
    L(G'_D) & L(T'_D) \arrow[u, "\semantic"'] \arrow[l, "\yield"]
  \end{tikzcd}
  \caption{The grammars considered, and their relation to each other.}
  \label{fig:grammar-square}
\end{figure}

@@

\noindent This ambiguity can be resolved however, since $\parse($\verb|'1 + (2 + 3)'|$) = \{t_1\}$ and $\parse($\verb|'(1 + 2) + 3'|$) = \{t_2\}$. To demonstrate the unresolvable case, we add the production $E -> s: E$ \verb|';'| $E$, at which point we find that the word \verb|'[1 ; 2]'| is unresolvably ambiguous; $\parse($\verb|'[1 ; 2]'|$) = \{t_3, t_4\}$ where:

@@

\noindent In this case, $t_4$ has an unambigous word (namely \verb|'[(1 ; 2)]'|), but $t_3$ does not. The solution is to modify the language definition in Figure~\ref{fig:running-example-definition} so that both non-terminals in the production $l$ are marked with $s$ (i.e., they look like $E_{\{s\}}$), at which point $\parse($\verb|'[1 ; 2]'|$) = \{t_1\}$.

We can expand this to an entire language, stating that a language is resolvably ambiguous iff all its words are resolvably ambiguous, alternatively:

@@

\noindent Additionally, on the dynamic side, there is another property of great practical relevance; we want ambiguities to be minimal and independent. In a practical setting, it is rarely helpful to merely state that an entire program is ambiguous. Instead, we wish to minimize the reported ambiguous portion of the program to, e.g., a sigle expression. Section~\ref{sec:local} addresses this.

@@

This section describes a decision algorithm for detecting unresolvable ambiguities in a language definition, starting with a version with several limitations, most of which are later lifted. The overarching goal is to find a tree $t \in L(G_t)$ such that there is no $w' \in L(G'_w)$ for which $\parse(w') = \{t\}$. Alternatively, find a tree that has no unambiguous words.

>>>>

\begin{figure}
\centering
\begin{minipage}{.47\textwidth}
  \centering
  \begin{tabular}{@{}ll@{}}
      Terminals & $t \in \T$ \\
      Non-terminals & $N \in \NT$ \\
      Labels & $l \in \Labels$ \\
      Marks & $m \subseteq \Labels$ \\
      Regular expressions & $r ::= t \mid N_m \mid r \cdot r $ \\
      & \hphantom{$r ::=$}$\mid r + r \mid \epsilon \mid r^{*}$ \\
      Labelled productions & $N -> l : r$ \\
  \end{tabular}
  \captionof{figure}{The abstract syntax of a language definition.}
  \label{fig:input-language-definition}
\end{minipage}\quad
\begin{minipage}{.45\textwidth}
  \centering
  \begin{tabular}{@{}l@{\quad$->$\quad}c@{ $:$\quad}l@{}}
    $E$ & $l$ & \verb|'['| ($E$ (\verb|';'| $E$)$^{*}$ $+$ $\epsilon$) \verb|']'| \\
    $E$ & $a$ & $E$ \verb|'+'| $E$ \\
    $E$ & $m$ & $E_{\{a\}}$ \verb|'*'| $E_{\{a\}}$ \\
    $E$ & $n$ & $N$ \\
  \end{tabular}
  \captionof{figure}{The input language definition used as a running example, an expression language with lists, addition, and multiplication, with precedence defined, but not associativity. Assumes that $N$ matches a numeric terminal.}
  \label{fig:running-example-definition}
\end{minipage}
\end{figure}

@@

\begin{figure}

  \begin{subfigure}[t]{.45\linewidth}
    \centering
    \begin{tabular}{@{}l@{\quad$->$\quad}c@{$($ }l@{}}
      \toprule
      $E$ & $l$ & \verb|'['| $(\epsilon + E($\verb|';'| $E)^{*})$ \verb|']'| $)$ \\
      $E$ & $a$ & $E$ \verb|'+'| $E$ $)$ \\
      $E$ & $m$ & $E$ \verb|'*'| $E$ $)$ \\
      $E$ & $n$ & $N$ $)$ \\
      \bottomrule
    \end{tabular}
    \caption{$T_D$, the parse trees of $G_D$.}
    \label{fig:running-example-generated:t}
  \end{subfigure}%
%
  \begin{subfigure}[t]{.45\linewidth}
    \centering
    \begin{tabular}{@{}l@{\quad$->$\quad}c@{$($ }l@{}}
      \toprule
      $E$ & $l$ & \verb|'['| $(\epsilon + E($\verb|';'| $E)^{*})$ \verb|']'| $)$ \\
      $E$ & $a$ & $E$ \verb|'+'| $E$ $)$ \\
      $E$ & $m$ & $E_{\{a\}}$ \verb|'*'| $E_{\{a\}}$ $)$ \\
      $E$ & $n$ & $N$ $)$ \\
      $E$ & $g$ & \verb|'('| $E$ \verb|')'| $)$ \\
      \midrule
      $E_{\{a\}}$ & $l$ & \verb|'['| $(\epsilon + E($\verb|';'| $E)^{*})$ \verb|']'| $)$ \\
      $E_{\{a\}}$ & $m$ & $E_{\{a\}}$ \verb|'*'| $E_{\{a\}}$ $)$ \\
      $E_{\{a\}}$ & $n$ & $N$ $)$ \\
      $E_{\{a\}}$ & $g$ & \verb|'('| $E$ \verb|')'| $)$ \\
      \bottomrule
    \end{tabular}
    \caption{$T'_D$, the parse trees of $G'_D$.}
    \label{fig:running-example-generated:t-prime}
  \end{subfigure}

  \begin{subfigure}[t]{.45\linewidth}
    \centering
    \begin{tabular}{@{}l@{\quad$->$\quad}l@{}}
      \toprule
      $E$ & \verb|'['| $E_{l1}$ \verb|']'| \\
      $E$ & $E$ \verb|'+'| $E$ \\
      $E$ & $E$ \verb|'*'| $E$ \\
      $E$ & $N$ \\
      \midrule
      $E_{l1}$ & $\epsilon$ \\
      $E_{l1}$ & $E$ $E_{l2}$ \\
      \midrule
      $E_{l2}$ & $\epsilon$ \\
      $E_{l2}$ & \verb|';'| $E$ $E_{l2}$ \\
      \bottomrule
    \end{tabular}
    \caption{$G_D$, the generated abstract syntax.}
    \label{fig:running-example-generated:w}
  \end{subfigure}%
%
  \begin{subfigure}[t]{.45\linewidth}
    \centering
    \begin{tabular}{@{}l@{\quad$->$\quad}l@{}}
      \toprule
      $E$ & \verb|'['| $E_{l1}$ \verb|']'| \\
      $E$ & $E$ \verb|'+'| $E$ \\
      $E$ & $E_{\{a\}}$ \verb|'*'| $E_{\{a\}}$ \\
      $E$ & $N$ \\
      $E$ & \verb|'('| $E$ \verb|')'| \\
      \midrule
      $E_{\{a\}}$ & \verb|'['| $E_{l1}$ \verb|']'| \\
      $E_{\{a\}}$ & $E_{\{a\}}$ \verb|'*'| $E_{\{a\}}$ \\
      $E_{\{a\}}$ & $N$ \\
      $E_{\{a\}}$ & \verb|'('| $E$ \verb|')'| \\
      \midrule
      $E_{l1}$ & $\epsilon$ \\
      $E_{l1}$ & $E$ $E_{l2}$ \\
      \midrule
      $E_{l2}$ & $\epsilon$ \\
      $E_{l2}$ & \verb|';'| $E$ $E_{l2}$ \\
      \bottomrule
    \end{tabular}
    \caption{$G'_D$, the generated concrete syntax.}
    \label{fig:running-example-generated:w-prime}
  \end{subfigure}
  \caption{The generated grammars.}
  \label{fig:running-example-generated}
\end{figure}

@@

\begin{figure}
  \begin{subfigure}[b]{.45\linewidth}
    \begin{center}
    \Tree [.{$m$}
        [.{$a$}
          [.{$n$} \terminal{1} ]
          \terminal{+}
          [.{$n$} \terminal{2} ] ]
        \terminal{*}
        [.{$n$} \terminal{3} ] ]
    \end{center}
    \[m(a(n(\terminal{1}) \terminal{+} n(\terminal{2})) \terminal{*} n(\terminal{3}))\]
    \caption{Example tree in $L(T_D)$ and visualization.}
  \end{subfigure}
  \begin{subfigure}[b]{.45\linewidth}
    \begin{center}
    \Tree [.{$m$}
        [.{$g$}
          \terminal{(}
          [.{$a$}
            [.{$n$} \terminal{1} ]
            \terminal{+}
            [.{$n$} \terminal{2} ] ]
          \terminal{)} ]
        \terminal{*}
        [.{$n$} \terminal{3} ] ]
    \end{center}
    \[m(g( \terminal{(} a(n( \terminal{1} ) \terminal{+} n( \terminal{2} )) \terminal{)} ) \terminal{*} n( \terminal{3} ))\]
    \caption{Example tree in $L(T'_D)$ and visualization.}
  \end{subfigure}

  \begin{subfigure}[b]{.40\linewidth}
    \[1 + 2 * 3\]
    \caption{Example word in $L(G_D)$.}
  \end{subfigure}
  \begin{subfigure}[b]{.40\linewidth}
    \[(1 + 2) * 3\]
    \caption{Example word in $L(G'_D)$.}
  \end{subfigure}
  \caption{Example with elements from each generated language that correspond to each other. The leaf terminals in the tree languages appear underlined to distinguish the two kinds of parentheses.}
  \label{fig:example-words-in-square}
\end{figure}
}

@@

\begin{figure}
  \begin{tikzcd}
    L(T_D) \arrow[d, "\yield"] & L(T'_D) \arrow[l, "\semantic"'] \arrow[d, "\yield"] \\
    L(G_D) & L(G'_D)
  \end{tikzcd}
  \caption{The grammars considered, and their relation to each other.}
  \label{fig:grammar-square}
\end{figure}

@@

\noindent This ambiguity can be resolved, since $\parse($\verb|'1 + (2 + 3)'|$) = \{t_1\}$ and $\parse($\verb|'(1 + 2) + 3'|$) = \{t_2\}$. To demonstrate the unresolvable case, we add the production $E -> s: E$ \verb|';'| $E$, at which point we find that the word \verb|'[1 ; 2]'| is unresolvably ambiguous; $\parse($\verb|'[1 ; 2]'|$) = \{t_3, t_4\}$ where:

@@

\noindent In this case, $t_4$ has an unambigous word (namely \verb|'[(1 ; 2)]'|), but $t_3$ does not. The solution is to modify the language definition in Figure~\ref{fig:running-example-definition} so that both non-terminals in the production $l$ are marked with $s$ (i.e., they look like $E_{\{s\}}$), at which point $\parse($\verb|'[1 ; 2]'|$) = \{t_3\}$.

We can expand this to an entire language, stating that a language is resolvably ambiguous iff all its words are resolvably ambiguous, or more concisely:

@@

\noindent Additionally, on the dynamic side, there is another property of great practical relevance; we want ambiguities to be minimal and independent. In a practical setting, it is rarely helpful to merely state that an entire program is ambiguous. Instead, we wish to minimize the reported ambiguous portion of the program to, e.g., a single expression. Section~\ref{sec:local} addresses this.

@@

This section describes a decision algorithm for detecting unresolvable ambiguities in a language definition, starting with a version with several limitations, most of which are later lifted. The overarching goal is to find a tree $t \in L(G_t)$ such that there is no $w' \in L(G'_w)$ for which $\parse(w') = \{t\}$, or prove that no such tree exists. Alternatively, find a tree that has no unambiguous words.
